Value-Suppressing Uncertainty Maps

### 2017-12-07 paper/main.tex
First typo pass.

We describe the motivations for VSUPs, provide examples of their utility for decision-making under uncertainty, and assess VSUPs in a crowdsourced experiment. Our results indicate that VSUPs create close integration between uncertainty and data, resulting in a demonstrably effect of decision-making, leading viewers to make more cautious choices that avoid high uncertainty.

@@

Given that bivariate maps, for reasons of either practicality or perceptual fidelity, have only a limited budget of outputs, we adapt an insight from Dunn~\cite{dunn1989dynamic} that these limited categories ought to be assigned with regards to their importance to analysis tasks, rather than uniformly. Our Value-Suppressing Uncertainty Palettes apply this insight to data with uncertainty. Frequently, uncertain information ought to be weighted less heavily in a decision-making process, or at least have less prominence in an analysis, than certain information.

@@

VSUPs rely on an underlying \textbf{quantization tree} that governs how values are discretized. Values about a certain uncertainty threshold are mapped to a singular ``root'' node. As uncertainty decreases, the tree branches into leaves, which evenly divide the data domain. The data value of a parent is the midpoint of all of its children. These leaves can then branch again, up to a designer-specified stopping point. The layers quantize the uncertainty domain, and sibling nodes quantize the data domain. \figref{fig:example} shows a VSUP with a tree with a branching factor of 2 and 4 layers, resulting in an output range with 15 distinct colors.

@@

By providing more color categories when the margin of error was relatively small, the VSUP map shows that, in many of the states leaning towards Hillary Clinton, her lead was quite small. The VSUP map, by limiting inferences about which candidate was leading when the uncertainty is high, also prevents the viewer from making strong claims about the outcome of the election, even if a state is leaning one way or the other. Given the relative volatility in polling data, this caution was warranted. An additional benefit to the VSUP map is that, as the diverging blue-red color map passes through a central whitepoint, it can be difficult to distinguish between uncertainty levels (which also interpolates to white) in the 2D map. The VSUP's tree-like structure separates these problematic colors.

@@

The results support our first hypothesis. Juxtaposition was a significant effect ($F(1,166)=5.5$,$p=0.02$). A post-hoc t-test confirmed ($t=2.5$, $p=0.01$) that superimposed charts ($M=0.58$, $SD=0.5$) performed significantly better than juxtaposed ($M=0.51$, $SD=0.49$) ones. For our identification task, juxtaposition requires at least two search tasks (one to locate squares with the correct data value, another to locate ones with the correct uncertainty value), which introduces additional error compared to conditions where value and uncertainty are displayed simultaneously. 
%This is reflected also in response times, which, although highly variable, were higher in juxtaposed conditions ($M=19$s, $SD=65$), than the other conditions ($M=13$s, $SD=38$), although this difference was not significant ($t=1.64$, $p=0.1$).

@@

\sizeFig

@@

VSUPs assume that uncertain values ought to have less visual impact that highly certain values. VSUPs therefore act both as a kind of filtering device as well as a bivariate representation. This assumption holds in scenarios like the virology dataset (\figref{fig:viralVsum}), where the expectation is that highly uncertain data is unreliable or should otherwise be downweighted in analysis. If the analyst has a different interpretation of uncertainty, or wishes to quickly and orthogonally analyze the distributions of uncertainty and value, other strategies, such as juxtaposed maps, may be more appropriate. VSUPs are designed for the \emph{integration} of value and uncertainty. Designers should take care in considering when and how this integration is desirable.

Our experiments dealt with cases where both uncertainty and value were represented by color. The perceptual non-separability of color channels is well-known~\cite{garner1970integrality, ware2012information}, and so the concept of a limited ``budget'' of distinguishable marks easier to quantify and illustrate. In principle, a VSUP can be created for any combination of visual variables. All that is required is a perceptual model of the interaction between these two variables. Where these models exist, as with the interaction between size and color~\cite{stone2014engineering}, the creation of VSUPs is straightforward. Where these models do not exist, or where the perceptual interaction is too complex to efficiently model, experimental work remains to be done before VSUPs can be considered a feasible design strategy. We are currently experimenting with the creation of VSUPs using these other channels, as in \figref{fig:size}.

>>>>

We describe the motivations for VSUPs, provide examples of their utility for decision-making under uncertainty, and assess VSUPs in a crowdsourced experiment. Our results indicate that VSUPs create close integration between uncertainty and data. This integration has a demonstrable effect on decision-making, leading viewers to make more cautious choices that avoid high uncertainty.

@@

Given that bivariate maps, for reasons of either practicality or perceptual fidelity, have only a limited budget of outputs, we adapt an insight from Dunn~\cite{dunn1989dynamic} that these limited categories ought to be assigned with regards to their importance to analysis tasks, rather than uniformly. Our Value-Suppressing Uncertainty Palettes apply this insight to data with uncertainty. Frequently, uncertain information ought to be weighted less heavily in a decision-making process, or at least have less prominence in an analysis, than information with high certainty.

@@

VSUPs rely on an underlying \textbf{quantization tree} that governs how values are discretized. Values above a certain uncertainty threshold are mapped to a singular ``root'' node. As uncertainty decreases, the tree branches into leaves, which evenly divide the data domain. The data value of a parent is the midpoint of all of its children. These leaves can then branch again, up to a designer-specified stopping point. The layers quantize the uncertainty domain, and sibling nodes quantize the data domain. \figref{fig:example} shows a VSUP with a tree with a branching factor of 2 and 4 layers, resulting in an output range with 15 distinct colors.

@@

By providing more color categories when the margin of error was relatively small, the VSUP map shows that, in many of the states leaning towards Hillary Clinton, her lead was quite small. The VSUP map, by limiting inferences about which candidate was leading when the uncertainty is high, also prevents the viewer from making strong claims about the outcome of the election in highly volatile states, even if a state is leaning one way or the other. Given the relative volatility in polling data, this caution is warranted. An additional benefit to the VSUP map is that, as the diverging blue-red color map passes through a central whitepoint, it can be difficult to distinguish between uncertainty levels (which also interpolates to white) in the 2D map. The VSUP's tree-like structure separates these problematic colors: the closest colors in the 2D scale are 2.9 units apart in CIELab space, but the closest colors are 6.2 units apart in the VSUP scale.

@@

The results support our first hypothesis. Juxtaposition was a significant effect ($F(1,166)=5.5$,$p=0.02$). A post-hoc t-test confirmed ($t=2.5$, $p=0.01$) that superimposed charts ($M=0.58$, $SD=0.50$) performed significantly better than juxtaposed ($M=0.51$, $SD=0.49$) ones. For our identification task, juxtaposition requires at least two search tasks (one to locate squares with the correct data value, another to locate ones with the correct uncertainty value), which introduces additional error compared to conditions where value and uncertainty are displayed simultaneously. 
%This is reflected also in response times, which, although highly variable, were higher in juxtaposed conditions ($M=19$s, $SD=65$), than the other conditions ($M=13$s, $SD=38$), although this difference was not significant ($t=1.64$, $p=0.1$).

@@

%\sizeFig

@@

VSUPs assume that uncertain values ought to have less visual impact that highly certain values. VSUPs therefore act both as a kind of filtering device as well as a bivariate representation. This assumption holds in scenarios like the virology dataset (\figref{fig:pollingVsum]}), where the expectation is that highly uncertain data is unreliable or should otherwise be downweighted in analysis. If the analyst has a different interpretation of uncertainty, or wishes to quickly and orthogonally analyze the distributions of uncertainty and value, other strategies, such as juxtaposed maps, may be more appropriate. VSUPs are designed for the \emph{integration} of value and uncertainty. Designers should take care in considering when and how this integration is desirable.

Our experiments dealt with cases where both uncertainty and value were represented by color. The perceptual non-separability of color channels is well-known~\cite{garner1970integrality, ware2012information}, and so the concept of a limited ``budget'' of distinguishable marks easier to quantify and illustrate. In principle, a VSUP can be created for any combination of visual variables. All that is required is a perceptual model of the interaction between these two variables. Where these models exist, as with the interaction between size and color~\cite{stone2014engineering}, the creation of VSUPs is straightforward. Where these models do not exist, or where the perceptual interaction is too complex to efficiently model, experimental work remains to be done before VSUPs can be considered a feasible design strategy. We are currently experimenting with the creation of VSUPs using these other channels, such as color and size.
